Modelling last glacial cycle ice dynamics in the Alps

### 2017-09-12 articles/cycle/alps-cycle.tex
Draft discussion of timing of the LGM.

    The modelled flow pattern at 24.56\,ka is complex (Fig.~\ref{fig:lgmvel}a).
    The modelled fast-flow regions generally occur along the main river
    valleys, while ice domes and ice divides are predominantly located over
    major reliefs . Nevertheless, the model results depict occasional ice flow
    across mountain passes (transfluences) and ice divide above topographic
    lows.

@@

    \begin{itemize}
    \item Maximum extent is time-transgressive (Fig.~\ref{fig:timing}).
    \item Glaciers have different response times (Fig.~\ref{fig:timing}).
    \end{itemize}

>>>>

    The modelled flow pattern at 24.57\,ka is complex (Fig.~\ref{fig:lgmvel}a).
    The modelled fast-flow regions generally occur along the main river
    valleys, while ice domes and ice divides are predominantly located over
    major reliefs . Nevertheless, the model results depict occasional ice flow
    across mountain passes (transfluences) and ice divide above topographic
    lows.

@@

    The timing of the LGM has been documented by radiocarbon and cosmogenic
    isotope dating techniques at multiple glacier lobes around the Alps.
    These data indicate that Alpine glaciers reached their maximum extent
    about synchronously with the Laurentide ice sheet around 25\,ka, but
    glacier lobes stayed in the foreland until as late as 20\,ka, a time
    around which the Fennoscandian ice sheet reached is maximum, while the
    Laurentide ice sheet was already in decay.

    In our simulation, the maximum areal cover is reached at 24.57\,ka
    (Fig.~\ref{fig:lgmvel}a), yet individual glacier lobes reach their
    maximum extent at different ages (Fig.~\ref{fig:timing}). For instance, the
    Dora Riparia, Ivrea, and Tagliamento glaciers reach an early maximum before
    27\,ka, the Durance, Rhone, Inn, Enns, Ticino, and Garda glaciers reach
    their maximum thickness in phase with the global Alpine areal maximum
    around 25\,ka, while the Isère, Como and Iseo glaciers reach a late maximum
    after 24\,ka (Fig.~\ref{fig:timing}). Remarkably, peripheral ice caps on
    the Vercors, Jura Mountains, Black Forest and Hochschwab reach their
    maximum extent even later, locally after 21\,ka.

    It is important to notice that these differences in timing of the LGM occur
    in the model results despite the simple climate forcing used consisting of
    homogeneous anomalies. These discrepancies between different glacier lobes
    are thus not related to climate, but are a pure product of modelled
    inherent glacier dynamics. They occur because of differences in the
    subglacial topography, in particular catchment sizes and bed roughness.

    PISM is a polythermal ice sheet model, i.e. it computes the distribution
    of englacial temperatures resulting from heat advection, vertical
    diffusion in ice and rock, and strain heat release, in three dimensions.
    For large Alpine glaciers flowing in deep valleys, such as the Rhone,
    Rhine and Garda glaciers, several thousands of years are needed to attain
    a thermodynamical equilibrium between cold ice advection from the high
    accumulation areas, upward diffusion of geothermal heat, and heat release
    from strong basal shear strain. Temperature evolution is, in part, limited
    by the slow warming of subglacial bedrock, that has been previously cooled
    downed by subfreezing air temperature before glacier advance.

    For such glaciers, this thermodynamical equilibrium is typically not yet
    attained around 25\,ka. This is why several Alpine lobes overshoot their
    balanced extent before thinning and receding towards the mountains as they
    warm towards thermodynamical equilibrium.

    Our results indicate that even for a relatively small ice sheet like that
    found in the Alps, differences in timing on the order of the millenium can
    result from complex glacier dynamics. Further differences in timing of the
    LGM caused by heterogeneous climatic anomalies, such as precipitation
    anomaly gradients, not included in our model set-up, may have
    counterbalanced or emphasize those presented here.
